---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
amd <- read.csv(unz("activity.zip", "activity.csv"))
gc()
```
Here is what the data looks like:

```{r}
head(amd)
str(amd)
```

The date variable needs to be converted into a date format.  
```{r}
amd$date <- as.Date(amd$date, "%Y-%m-%d")

##Verify that reformatting was successful

class(amd$date)
```

Lastly, group the data by date and time interval, using the group_by() function in the dplyr package. 

```{r}
library(dplyr)
amd <- group_by(amd, date)
```

## What is mean total number of steps taken per day?

Using the summarize function from dplyr.  

```{r}
totalsteps <- summarize(amd, sum=sum(steps))
```

Here is a histogram, which shows the distribution of the number of steps taken per day:

```{r}
library(ggplot2)
ggplot() + aes(na.omit(totalsteps$sum)) + geom_histogram(binwidth = 1000, color = "black", fill = "blue")
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
