---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, I loaded the data in using read.csv().  The code assumes that the folder containing the dataset is your working directory, and is called "activity.csv".  Because it is a zip file, the unz() function is used to open a connection inside the zip file and extract the csv file.  The gc() function is the safest and easiest way to close the connection after the data is read in.   

```{r warning=FALSE}
amd <- read.csv(unz("activity.zip", "activity.csv"))
gc()
```
Here is an overview of what the data looks like:

```{r}
head(amd)
str(amd)
```

The date variable then needs to be converted into a date format, using the as.Date() function.    
```{r}
amd$date <- as.Date(amd$date, "%Y-%m-%d")

##Verify that reformatting was successful

class(amd$date)
```

Lastly, group the data by date and time interval, using the group_by() function in the dplyr package. Each type of grouping will be used later.  

```{r, message=FALSE}
library(dplyr)
grp_by_date <- group_by(amd, date)
grp_by_interval <- group_by(amd, interval)
```

## What is mean total number of steps taken per day?

Using the summarize function from dplyr, a new dataframe is created which gives the total number of steps on each date in the sample.   

```{r}
totalSteps <- summarize(grp_by_date, sum=sum(steps))
head(totalSteps)
```

This histogram shows the distribution of total number of steps that we just calculated:

```{r}
library(ggplot2)
ggplot() + aes(na.omit(totalSteps$sum)) + geom_histogram(binwidth = 1000, color = "black", fill = "blue") + labs(title = "Distribution of the Total Number of Steps ", x = "Total number of Steps")
```

The mean and median of the total steps taken per day can calculated fom the same dataset:

```{r}
mn <- mean(totalSteps$sum, na.rm = TRUE)
md <- median(totalSteps$sum, na.rm = TRUE)
mn
md
```


## What is the average daily activity pattern?

Again, using the summarize() function in dplyr, a new dataframe is created which gives the average number of steps recorded for each time interval:

```{r}
averageSteps <- summarize(grp_by_interval, mean = mean(steps, na.rm = TRUE))
```

Here is the time series plot by time interval, which represents the data we just calculated:

```{r}
ggplot() + aes(averageSteps$interval, averageSteps$mean) + geom_line(size = 1.5, color = "purple") + labs(title = "Time Series of Average Steps by 5-minute Interval", x = "Interval", y = "Average # of Steps")
```

The interval with the most number of steps is given in the following code:

```{r}
as.numeric(averageSteps[which.max(averageSteps$mean), 1])
```


## Imputing missing values

The original dataset has many missing values, which can potentially bias our analysis in unknown ways.  The number of missing values for number of steps is calculated in the following code:

```{r}
sum(is.na(amd$steps))
```

My strategy for dealing with missing values is to replace the NA value with the average number of steps for each time interval that was taken from the rest of the dataset. I believe this is the best way to do so:

```{r}

```


## Are there differences in activity patterns between weekdays and weekends?
